---
title: "Time Series Analysis in R"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

### Background:

This repository is for the development and maintenance of R teaching material for use in the EPIET (European Programme of Intervention Epidemiology Training) and UK FETP (United Kingdom Field Epidemiology Training Programme) **Time Series Analysis (TSA) module**.

The module consists of a single computer-based practical exercise that uses multiple datasets to demonstrate different facets of time series analysis; the help guide for these exercises has been translated from the STATA version into R. 


### Acknowledgements:

This material is based on an equivalent practical guide for STATA. The guide is presented as an R markdown document; please see the title pages for copyright details and a list of authors and contributors.
 
 
### Requirements:

Each exercise requires the use of one or more teaching datasets, which are included in this repository.

To run the R markdown guide or R template, participants and facilitators will need to install the following software:

 - R (download and install the latest version from CRAN [here](https://cran.r-project.org/))
 - RStudio (download the latest version as an installer or ready-to-use `.zip` file [here](https://rstudio.com/products/rstudio/download/))
 - Rtools (download and install the latest version from CRAN [here](https://cran.r-project.org/bin/windows/Rtools/))

A list of required R packages and functions, as well as code to install them has been included in the first chunk of each R markdown document.


### How to use:

After installing the above software, clone this repository either by clicking on the green `Clone or download` button on this page, or by entering the following command into git bash:

```{r, eval=FALSE}
git clone https://github.com/EPIET/TimeSeriesAnalysis.git
```


Then:

 - Open the .Rproj file in RStudio
 - Navigate to the `Files` tab within RStudio 
 - Click on the relevant R markdown document to open it
 
To provide a printed copy of the R markdown practical guide for participants:

 - Open the relevant R markdown document in RStudio (as above)
 - Click on the `Knit` button and select `Knit to pdf`
 - This will save a .pdf version of the document in your working directory, which can then be printed.


### Maintenance:

This project is currently being maintained by [Amy Mikhail](https://github.com/AmyMikhail).  

Contributions are welcome: please contact the maintainer to request access.

To report bugs or make feature requests, please post an issue [here](https://github.com/EPIET/TimeSeriesAnalysis/issues).
